A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-02 tex/codatadp.tex
Refactoring

First, Figure~\ref{fig:cc-rules} gives the rules for constructing the bidirectional closure of $\Ec$.
The rules \rn{Refl}, \rn{Symm}, \rn{Trans}, and \rn{Cong} together compute the (upwards) congruence closure,
after which a conflict may be recognized by \rn{Conflict} when an inconsistency is found.
The rules \rn{Unify$_1$} and \rn{Unify$_2$} together compute (downwards) unification,
where additional equalities are inferred based on the injectivity of constructors by \rn{Unify$_1$},
and failures to unify equivalent terms are recognized by \rn{Unify$_2$}.

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Cyclic}
\(
\inferrule{
  t : \tau \text{ inductive } 
  \quad
  p \in \ttpath{\Ec}{\ec{t}}{\ec{t}}
}{
  \bot
}
\)
\\[3.7ex]
\rn{Bisimilar}
\(
\inferrule{
 t_1, t_2 : \tau \text{ co-inductive }
  \quad
 \exists \sigma. p \sigma \in \tpath{\Ec}{\ec{t_1}} \text{ for each } p \in \tpath{\Ec}{\ec{t_2}}
}{
 \Ec := \Ec, t_1 \teq t_2
}
\)
\end{tabular}
\caption{Rules for cyclicity and bisimilarity.
}
\label{fig:ab-rules}
\end{figure}

@@

\rem{Change notation for inductive/co-inductive.}

@@

\begin{lemma}[Termination]
All derivation trees are finite.
\end{lemma}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S^0$ be the set of terms that are arguments of a selector application in $\tEc$.
For $i \geq 0$, let $S^{i+1}$ $=$ $S_i \cup \{ s( t ) \mid t \in S^i, s \in sel( t ), t : \tau \text{ finite} \}$.
Let $S^\ast$ be the fix point of this sequence, which is guarenteed to be finite 
since $S^\ast \setminus S^0$ only includes selectors applied to terms having finite type.
Now, let $\tcEc$ be the set all subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_j}_i( t ) ) \mid t \in S^\ast, C_i \in cons( t ) \}$.
In a derivation tree with root node $\Ec$, 
it can be shown by case analysis on the rules of the calculus that each non-root node is 
obtained as a result of adding an equality between two terms from $\tcEc$ to its parent node.
Thus, the maximum depth of a branch in a derivation tree with root node $\Ec$ is $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\end{proof}

\begin{lemma}[Refutation Soundness]
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
TODO
\end{proof}

>>>>

Figure~\ref{fig:cc-rules} gives the rules for constructing the bidirectional closure of $\Ec$.
The rules \rn{Refl}, \rn{Symm}, \rn{Trans}, and \rn{Cong} together compute the (upwards) congruence closure,
after which a conflict may be recognized by \rn{Conflict} when an inconsistency is found.
The rules \rn{Unify$_1$} and \rn{Unify$_2$} together compute (downwards) unification,
where additional equalities are inferred based on the injectivity of constructors by \rn{Unify$_1$},
and failures to unify equivalent terms are recognized by \rn{Unify$_2$}.

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Cyclic}
\(
\inferrule{
  t : \tau, \tau \in \Data
  \quad
  p \in \ttpath{\Ec}{\ec{t}}{\ec{t}}
}{
  \bot
}
\)
\\[3.7ex]
\rn{Bisimilar}
\(
\inferrule{
 t_1, t_2 : \tau, \tau \in \Codata
  \quad
 \exists \sigma. p \sigma \in \tpath{\Ec}{\ec{t_1}} \text{ for each } p \in \tpath{\Ec}{\ec{t_2}}
}{
 \Ec := \Ec, t_1 \teq t_2
}
\)
\end{tabular}
\caption{Rules for cyclicity and bisimilarity.
}
\label{fig:ab-rules}
\end{figure}

@@

@@

\begin{lemma}[Termination]
All derivation trees are finite.
\end{lemma}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S^0$ be the set of terms that are arguments of a selector application in $\tEc$.
For $i \geq 0$, let $S^{i+1}$ $=$ $S_i \cup \{ s( t ) \mid t \in S^i, s \in sel( t ), t : \tau \text{ finite} \}$.
Let $S^\ast$ be the fix point of this sequence, which is guarenteed to be finite 
since $S^i \setminus S^0$ only includes selectors applied to terms having finite type.
Now, let $\tcEc$ be the set of subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_j}_i( t ) ) \mid t \in S^\ast, C_i \in cons( t ) \}$.
In a derivation tree with root node $\Ec$, 
it can be shown by case analysis on the rules of the calculus that each non-root node is 
obtained as a result of adding an equality between two terms from $\tcEc$ to its parent node.
Thus, the maximum depth of a branch in a derivation tree with root node $\Ec$ is $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\end{proof}

\begin{lemma}[Refutation Soundness]
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
%We show that if a derivation tree with $\Ec$ has a model $\M$, 
\end{proof}
